A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-30 paint-contour-tree.tex
figs updated
updates to intro

For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical point},
exactly where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
representing the important topological events. An edge connects two critical points if

Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}.

@@

@@

We introduce the concept of path decompositions. A \emph{leaf path} is simply a contiguous path in $T$
containing a leaf, which is monotonic in the height values of its vertices.
% Then in the following two subsections 
%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

\begin{definition}
\label{def:path} A \emph{path decomposition}, $P(T)$, is a partition of $T$ into 
 a set of vertex disjoint leaf paths.  
% 
% The \emph{cost} of $P(T)$ is $\cost(P(T)) = \sum_{p\in P(T)} |p| \log |p|$. (We use $|p|$ to denote the number of vertices in $p$.)
\end{definition}

>>>>

For convenience, imagine $f$ to be smooth. All points where the gradient of $f$ is zero are \emph{critical point},
exactly where the topology of the contours change. The vertices of the contour tree are exactly these critical points,
representing the important topological events. An edge connects two critical points if one event immediately ``follows"
the other as the sweep plane makes its pass. (We provide formal definitions later.)

\Fig{relative} and \Fig{sorting} show examples of simplicial complices with heights and their contour trees.
Think of the edges as point downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such point. Leaves of up-degree
$1$ are minimum, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits". In these examples ($d=2$), contours are closed loops.
In dimensions $d \geq 3$, we may have internal
vertices of both up and down-degree $1$. The correspond to contours (which are themselves manifolds) gaining or losing handles.

Suppose we have $f: \MM \mapsto \RR$, where $\MM$ has $n$ vertices, $N$ facets in total, and $t \leq n$ critical points.
It is standard to assume that $f$ is PL-Morse (a well-behaved assumption) and that the maximum degree in the contour tree
is $3$. A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N)$ time~\cite{csa-cctad-00}. (Henceforth, $\alpha(\cdot)$ denotes the inverse
Ackermann function.) One cannot overstate the importance of this result, since 
subsequent practical and theoretical work in contour trees uses their ideas. It is natural to think of $N$ as $\Theta(n)$
in practical applications (certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang et al build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely $\Omega(n)$.
So we are still left with the bottleneck of sorting a large set of numbers. Unfortunately, there is a worst-case
lower bound of $\Omega(t\log t)$ from Chiang et al~\cite{cllr-sooscctmp-05}, based on a construction of Bajaj et al~\cite{BaKr+98}.

But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.

In a contour tree $T$, a \emph{leaf path} is simply a contiguous path in $T$
containing a leaf, which is also monotonic in the height values of its vertices. A leaf path runs either upwards
or downwards, never both.
% Then in the following two subsections 
%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

\begin{definition}
\label{def:path} For a contour tree $T$, a \emph{path decomposition}, $P(T)$, is a partition of $T$ into a set of vertex disjoint leaf paths.  
\end{definition}

\begin{theorem} \label{thm:alg-main} Consider $f:\MM \mapsto \RR$ that is PL-Morse, where the contour tree $T$
has maximum degree\footnote{The maximum degree $3$ assumption is primarily there to simplify the presentation and proof. Our algorithm
does not need the assumption.} $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implictly by the algorithm).
\end{theorem}

Let us interpret the running time. 
For any path decomposition $P(T)$ and any $p \in P(T)$, obviously $|p| \leq |T| = t$.
So the running time is clearly $O(t\log t + N)$. But this is only the case when $T$ is tall and thin.  
An obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
Our running time is now $O(t\alpha(t))$, a significant improvement from $O(t\log t)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{alg-main}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

@@

Contour trees were (unsurprisingly) used to study terrain maps by Boyell and Ruston, and Freeman and Morse~\cite{BoRu63,FrMo67}.
They were used for isoline extraction in geometric data by van Krevald et al~\cite{kobps-ctsssit-97}, who provided
the first formal algorithm. 
Contour trees and related topological data structures have been applied in analysis of fluid mixing, combustion simulations,
and studying chemical systems~\cite{LaBe+06,BrWe+10,BeWe+11,BrWe+11,MaGr+11}. Carr's thesis gives various
applications of contour trees for data visualization~\cite{c-tmi-04}.

@@
